Value-Suppressing Uncertainty Maps

### 2017-03-23 paper/figures.tex
First draft of related works section
Added performance figure.

\newcommand{\exampleFig}{
\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{example.pdf}
	\caption{A comparison between a standard bivariate map and a VSUM. Both use the same visual channels to encode value (position along the Viridis color map) and uncertainty (lightness and saturation), and have an equal standard of perceptual discriminability (at least 18 units of distance in CIELAB color space between colors). However, highly uncertain values result in colors that are very close together in the bivariate case, meaning the full bivariate map is only 9 bins under these constraints--- a 4x4 map would result in colors that are perceptually too close together. By contrast, the VSUM intentionally reduces bins when uncertainty is high, eventually aliasing all highly uncertain values to the same color. This decision affords more distinct colors in other regions of the map, and so an increase in overall bins to 15. The resulting map suppresses the value of uncertain data, but increases the discriminability of highly certain data.}
	\label{fig:example}
\end{figure}

}

>>>>

\newcommand{\exampleFig}{
\begin{figure}
	\centering
	\includegraphics[width=0.9\columnwidth]{example.pdf}
	\caption{A comparison between a standard bivariate map and a VSUM. Both use the same visual channels to encode value (position along the Viridis color map) and uncertainty (lightness and saturation), and have an equal standard of perceptual discriminability (at least 18 units of distance in CIELAB color space between colors). However, highly uncertain values result in colors that are very close together in the bivariate case, meaning the full bivariate map is only 9 bins under these constraints--- a 4x4 map would result in colors that are perceptually too close together. By contrast, the VSUM intentionally reduces bins when uncertainty is high, eventually aliasing all highly uncertain values to the same color. This decision affords more distinct colors in other regions of the map, and so an increase in overall bins to 15. The resulting map suppresses the value of uncertain data, but increases the discriminability of certain data.}
	\label{fig:example}
\end{figure}
}

\newcommand{\performanceFig}{
	\begin{figure}
		\centering
		\includegraphics[width=0.9\columnwidth]{performance.pdf}
		\caption{
			The number of discrete color categories in traditional bivariate color maps versus VSUMs, assuming value is encoded by the Viridis color map~\protect\cite{garnier2015viridis}, and uncertainty is encoded by saturation/value (whiter values are more uncertain). Since uncertain colors are closer together than certain colors, and VSUMs intentionally require fewer color bins in those regions, VSUMs always have more color categories than the corresponding 2D bivariate map. The red lines denote:\\
		    \textbf{1)} Theoretical 50\% JND from CIELAB specification. \\
			\textbf{2)} 50\% JND from controlled lab study of Mahy et al.~\protect\cite{mahy1994evaluation}.\\
			\textbf{3)} 50\% JND from Szafir et al.\protect\cite{szafir2014adapting}.\\
			\textbf{4)} 50\% JND for small visual targets from Stone et al.~\protect\cite{stone2014engineering}.\\
			\textbf{5)} 50\% The threshold resulting in the simplest possible (2x2) bivariate map. \\
	    }
		\label{fig:example}
	\end{figure}
}

### 2017-03-23 paper/template.tex
Typo in related works
Added performance figure.

Superposition, or the overlay of glyphs representing uncertainty on top of a data map, is other common strategy.  Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe et al.~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of this technique, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying datamap (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions (for instance, simultaneous contrast effects) that can impair the legibility of glyphs, and the necessity of performing spatial binning in order to ensure that there are a discrete number of uncertainty glyphs.

@@

Correll et al. use a version of VSUMs in their LayerCake genomics visualization tool \cite{correll2015layercake,correll2011visualizing}, where the reads with insufficient certainty are mapped to a smaller and smaller set of increasing grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent.

>>>>

Superposition, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy.  Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe et al.~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of this technique, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying datamap (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions (for instance, simultaneous contrast effects) that can impair the legibility of glyphs, and the necessity of performing spatial binning in order to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, and suppress important signals, or highlight spurious ones \cite{battersby2016shapes}.

@@

Correll et al. use a version of VSUMs in their LayerCake genomics visualization tool \cite{correll2015layercake,correll2011visualizing}, where marks representing genomic data with insufficient certainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection ot uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both of these cases, this property of aliasing is ad hoc, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes this choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

\subsection{Design Considerations}

\performanceFig

There are multiple choices that designers must make before creating a VSUM. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a metric for evaluating discriminability), and 3) a degradation function to determine how quickly uncertain values are aliased. Often, the graphical conventions of a chart
